Acceleration of Bilateral filtering algorithm for manycore and multicore architectures

### 2012-03-11 LateX/paper8.tex
Updated text and arrangement

\documentclass{IEEEtran}
\usepackage{color}
\usepackage{array}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{paralist}
\usepackage{cite}
 \usepackage{color}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{pgf}
\usepackage{pgfplots}
\usepackage{url}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}

\begin{document}
\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive Bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

Our work explores efficient parallel implementation of Bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures. Bilateral filtering kernel is used to smooth an image without blurring its edges. It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.

A faster approximation to the original algorithm \cite{Paris2009} extends the image to 3D space by treating the pixel intensity as the third dimension and using a linear shift-invariant convolution in 3D. A separable implementation \cite{Pham2005} applies the filter kernel first to one of the image dimensions and the intermediate result is filtered again on other dimensions. This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$, where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image. Yang et al.\cite{yang2009} have reported a constant time algorithm by dividing the Bilateral filtering kernel into a constant number of time spatial kernels. However, we do not parallelize these methods since our interest is to develop a parallelization method that can be applied, in general, to any spatially invariant and inseparable filtering algorithm.

In this paper, we propose a set of optimizations and a non-naive algorithm for Bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half. We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions. We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.

@@

Rest of the paper is organized as following: Section~\ref{sec:stencil} defines the Bilateral filtering kernel and introduces a naive algorithm and our pair-symmetric algorithm for Bilateral filtering. Section~\ref{sec:GPUoptimizations} introduces the optimizations employed to devise the pair-symmetric algorithm for CUDA architectures. Section~\ref{sec:optimizations} introduces the optimizations for CPU architectures and establishes the rationale behind ignoring traditional memory related optimizations for Bilateral filtering kernel. Our hardware and software testbed for experiments is detailed in Section~\ref{sec:expt}, immediately followed by Section~\ref{sec:results} that reports the performance of Bilateral filtering kernel on both multicore and manycore architectures. Section~\ref{sec:conclusion} concludes this paper with remarks on general applicability of our optimizations.

@@

Bilateral filtering combines two different kernels - the spatial kernel and the photometric kernel. With the photometric kernel, when the Center of the Filter rests on a pixel that boarders or is a part of a pictorial edge, e.g., an outline of an object or person against a contrasting background, the Neighbors that make up the other (contrasting) side of the edge have an insignificant impact on the Center Pixel's corresponding output pixel, which is stored in the destination array storing the processed image. Furthermore, when the Center is not bordering, and is not a part of a pictorial edge, the Neighbors contribute equally and the noise is filtered out. Hence, pictorial edges maintain sharpness while noise in the pictorial regions are filtered, resulting in more appealing images. Mathematically, the Bilateral filter can be defined in terms of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:\\

@@

The combined kernel, which is the Bilateral filter, is the product
of spatial kernel and photometric kernel and is described as: 
\begin{equation}
h(x)=k^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)s(f(\xi),f(x))d\xi\label{combined}
\end{equation}

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the Bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:

\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

In addition to restating the Bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the Bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the Bilateral filter has two data dependencies. As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the Bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

\subsection{A Naive Approach to Bilateral filtering kernel}

\begin{figure}
\centering\includegraphics[scale=0.35]{images/the1}
\includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
\caption{Example of a naive Bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly used in Bilateral filtering kernel. Previous implementations\cite{Paris2009,Pham2005} of Bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken. Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup. Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive Bilateral filtering kernel.

%\begin{algorithm}[!h]
%\caption{Generalized Naive Stencil Kernel Algorithm}
%\label{alg:Example-Algorithm-float}
%\begin{algorithmic}
%\FORALL{element  $s_{i,j}$ in source image s}
%\FORALL{data dependency $n$}
%\STATE{$X_n \leftarrow 0$}
%\ENDFOR
%\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
%\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
%\FORALL{data dependency $n$}
%\STATE{$X_n \leftarrow X_n + f_n(s[i,j],s[i+k,j+l])$}
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\STATE{output[$i,j] \leftarrow f_{output}(X_{1}, \dots, X_{last})$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

\begin{algorithm}[!h]
\caption{Naive Bilateral Filter Algorithm}
\label{alg:Example-Algorithm-float}
\begin{algorithmic}
\FORALL{element  $s_{i,j}$ in source image s}
\STATE{$X_{1} \leftarrow 0$}
\STATE{$X_{2} \leftarrow 0$}
\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\ENDFOR
\ENDFOR
\STATE{output[$i,j] \leftarrow X_{1} / X_{2}$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Pair-symmetric Bilateral filtering kernel}
The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half. Thus, it can theoretically cut the processing time in half at the cost of increased memory requirement, which increases with the amount of dependencies. Its implementation is non-trivial due to potential race-conditions.

Our pair-symmetric Bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations. For example, in the Bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$. It then follows that the naive Bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.

The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive Bilateral filtering algorithm. Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors. The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$. This is trivial as the naive Bilateral filtering algorithm already does this. What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency. So, there are $n$ Maps for $n$ number of dependencies.

After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps. As applied to the Bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result. After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.

%\begin{algorithm}[h]
%\caption{Pair-symmetric Stencil Algorithm}
%\label{alg:Example-Algorithm-float-1}
%\begin{algorithmic}
%\STATE /* D is the set of dependency arrays */
%\FORALL{data dependency $n$}
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$D_n[i,j] \leftarrow 0$}
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image $s$}
%\FOR{$k=1$ \TO $filter\_hw$}
%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
%\FORALL{data dependency $n$}
%\STATE $temp \leftarrow f_n(s[i,j],s[i+k,j+l])$
%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
%\STATE $D_n[i+k,j+l] \leftarrow D_n[i+k,j+l] + temp$
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\FOR{$l = -filter\_hw$ \TO $0$}
%\FORALL{data dependency $n$}
%\STATE $temp \leftarrow f_n(s[i,j],s[i,j+l])$
%\STATE $D_n[i,j] \leftarrow D_n[i,j] + temp$
%\STATE $D_n[i,j+l] \leftarrow D_n[i,j+l] + temp$
%\ENDFOR
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$output[i,j] \leftarrow f_{output}(D_{1}[i,j], \dots, D_{last}[i,j])$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

%\begin{algorithm}[h]
%\caption{Pair-symmetric Bilateral Filtering kernel}
%\label{alg:Example-Algorithm-float-1}
%\begin{algorithmic}
%\STATE /* D is the set of dependency arrays */
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$D_1[i,j] \leftarrow 0$}
%\STATE{$D_2[i,j] \leftarrow 0$}
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image $s$}
%\FOR{$k=1$ \TO $filter\_hw$}
%\FOR{$l = -filter\_hw$ \TO $filter\_hw$}
%\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i+k,j+l])^{2}}$}
%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i+k,j+l]$}
%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
%\STATE{$D_1[i+k,j+l] \leftarrow D_1[i+k,j+l] + temp \cdot s[i,j]$}
%\STATE{$D_2[i+k,j+l] \leftarrow D_2[i+k,j+l] + temp$}
%\ENDFOR
%\ENDFOR
%\FOR{$l = -filter\_hw$ \TO $0$}
%\STATE{$temp \leftarrow e^{C_{1}l^2 + C_2(s[i,j] - s[i,j+l])^{2}}$}
%\STATE{$D_1[i,j] \leftarrow D_1[i,j] + temp \cdot s[i,j+l]$}
%\STATE{$D_2[i,j] \leftarrow D_2[i,j] + temp$}
%\STATE{$D_1[i,j+l] \leftarrow D_1[i,j+l] + temp \cdot s[i,j]$}
%\STATE{$D_2[i,j+l] \leftarrow D_2[i,j+l] + temp$}
%\ENDFOR
%\ENDFOR
%\FORALL{element $s_{i,j}$ in source image s}
%\STATE{$output[i,j] \leftarrow D_{1}[i,j] / D_{2}[i,j]$}
%\ENDFOR
%\end{algorithmic}
%\end{algorithm}

@@

\section{CUDA-based Implementation Details}
\label{sec:GPUoptimizations} 
As Bilateral filtering algorithm belongs to the class of embarrassingly parallel algorithms, CUDA is well suited for its implementation. However,
in order to effectively implement such an algorithm in CUDA, one must understand how CUDA operates, which is significantly different from any CPU.

@@

In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory. Each warp executes one instruction at a time for its own group of threads. Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls. Hence, synchronization between threads in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that meet the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*j_pitch + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls as shown in Figure~\ref{fig:pspro}. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.

\subsection{Shared Memory and Tile Processing}
\begin{figure}
\includegraphics[width = 0.48 \columnwidth]{images/project24.pdf}
\caption{Illustration of our CUDA implementation of pair-symmetric processing using implicit intra-wap synchronization for iterating through the pixel-cells for each thread-block's assigned tile.}
\label{fig:pspro}
\end{figure}

>>>>

\documentclass{IEEEtran}
\usepackage{color}
\usepackage{array}
\usepackage{float}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{paralist}
\usepackage{cite}
 \usepackage{color}
\usepackage{amsfonts}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage{tikz}
\usepackage{pgf}
\usepackage{pgfplots}
\usepackage{wrapfig}
\usepackage{url}
\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}

\begin{document}
\title{Acceleration of Bilateral filtering algorithm for manycore and multicore architectures}
\author{\IEEEauthorblockN{Dinesh Agarwal, Sami Wilf, Abinashi Dhungel, and Sushil K. Prasad}\\
\IEEEauthorblockA{ Georgia State University\\
 Department of Computer Science\\
 Atlanta, Georgia 30303\\
 }}
\maketitle
\begin{abstract}
This work explores multicore and manycore acceleration for the embarrassingly parallel, compute-intensive bilateral filtering kernel. For manycore architectures, we have created a pair-symmetric algorithm to avoid redundant calculations and for multicore architectures we improve the algorithm by use of low level single instruction multiple data (SIMD) parallelism across multiple threads. We present empirical data evidencing the performance gains we achieved over variety of hardware architectures including Nvidia GTX280, AMD Barcelona, AMD Shanghai, Intel Harpertown, AMD Phenom, Intel Core i7 quad core, and Intel Nehalem 32 core machines. The best speedup achieved was a 235.5x speedup by our CUDA-based implementation of our pair-symmetric algorithm running on Nvidia's GTX280. Our CPU multicore implementations resulted in a speedup of up to 38x using 16 cores of AMD Barcelona each with 4-stage vector pipeline, up to 23x using 8 cores for Intel Harpertown, and up to 26x using 32 cores of Intel Nehalem multicore machines compared to a compiler-optimized code.
\end{abstract}
% For peer review papers, you can put extra information on the cover
% page as needed:
% \ifCLASSOPTIONpeerreview
% \begin{center} \bfseries EDICS Category: 3-BBND \end{center}
% \fi
% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle
\section{Introduction}
\label{sec:intro} The recent advances in parallel computing and architecture design have provided a series of avenues for application performance improvements. However, unlike the days when applications automatically became faster with increased clock frequency, newer chips require programmers and researchers to revisit their algorithm designs to harness the power of underlying architectures. Moreover, there is no well defined list of guidelines that can be followed for each application, which makes it challenging to deploy existing applications, even fundamental ones, to a myriad of available multicore and manycore architectures.

Our work explores efficient parallel implementation of bilateral filtering kernel\footnotemark[1] proposed by Tomasi and Manduchi \cite{Tomasi1998}, on multicore and manycore architectures. bilateral filtering kernel is used to smooth an image without blurring its edges. It is ubiquitous in various image processing applications in various contexts such as denoising\cite{zhang2008}, texture editing and relighting\cite{eisemann2004}, tone management\cite{Bae2006}, stylization\cite{DeCarlo2002}, optical-flow estimations\cite{Xiao2006}, and demosaicking\cite{Ramanath2003} etc.

A faster approximation to the original algorithm \cite{Paris2009} extends the image to 3D space by treating the pixel intensity as the third dimension and using a linear shift-invariant convolution in 3D. A separable implementation \cite{Pham2005} applies the filter kernel first to one of the image dimensions and the intermediate result is filtered again on other dimensions. This reduces the complexity from $O(N*(m^{d}))$ to $O(N*d)$, where $N$ is the total number of pixels in the image and $d$ is the dimensions of the image. Yang et al.\cite{yang2009} have reported a constant time algorithm by dividing the bilateral filtering kernel into a constant number of time spatial kernels. However, we do not parallelize these methods since our interest is to develop a parallelization method that can be applied, in general, to any spatially invariant and inseparable filtering algorithm.

In this paper, we propose a set of optimizations and a non-naive algorithm for bilateral filtering kernel. Specifically, we introduce a pair-symmetric algorithm which has the theoretical potential to reduce processing time to half. We propose architecture specific optimizations, such as exploiting the unique capabilities of special registers available in modern multicore architectures and the rearrangement of data access patterns as per the computations to exploit special purpose instructions. We also propose optimizations pertinent to Nvidia's CUDA, including utilization of CUDA's implicit synchronization capability and the maximization of single-instruction-multiple-thread efficiency.

@@

Rest of the paper is organized as following: Section~\ref{sec:stencil} defines the bilateral filtering kernel and introduces a naive algorithm and our pair-symmetric algorithm for bilateral filtering. Section~\ref{sec:GPUoptimizations} introduces the optimizations employed to devise the pair-symmetric algorithm for CUDA architectures. Section~\ref{sec:optimizations} introduces the optimizations for CPU architectures and establishes the rationale behind ignoring traditional memory related optimizations for bilateral filtering kernel. Our hardware and software testbed for experiments is detailed in Section~\ref{sec:expt}, immediately followed by Section~\ref{sec:results} that reports the performance of bilateral filtering kernel on both multicore and manycore architectures. Section~\ref{sec:conclusion} concludes this paper with remarks on general applicability of our optimizations.

@@

Bilateral filtering combines two different kernels - the spatial kernel and the photometric kernel. With the photometric kernel, when the Center of the Filter rests on a pixel that boarders or is a part of a pictorial edge, e.g., an outline of an object or person against a contrasting background, the Neighbors that make up the other (contrasting) side of the edge have an insignificant impact on the Center Pixel's corresponding output pixel, which is stored in the destination array storing the processed image. Furthermore, when the Center is not bordering, and is not a part of a pictorial edge, the Neighbors contribute equally and the noise is filtered out. Hence, pictorial edges maintain sharpness while noise in the pictorial regions are filtered, resulting in more appealing images. Mathematically, the bilateral filter can be defined in terms of the spatial kernel and photometric kernel as follows\cite{Tomasi1998}:\\

@@

The combined kernel, which is the bilateral filter, is the product
of spatial kernel and photometric kernel and is described as: 
\begin{equation}
h(x)=k^{-1}(x)\int_{-\infty}^{\infty}\int_{-\infty}^{\infty}f(\xi)c(\xi,x)s(f(\xi),f(x))d\xi\label{combined}
\end{equation}

@@

Function $c(\xi,x)$ measures the $geometric~closeness$ between neighborhood Center $x$ and a Neighbor $\xi$. Function $s(f(\xi),f(x))$ measures the $photometric$ $similarity$ (similarity in pixel intensity) between the same. The influence of a Neighbor on the Filter's output computation, for a Center Pixel, is calculated as a product of spatial kernel and the photometric kernel. Thus, Neighbors with similar photometric properties influence the Center Pixel more than those with less similar photometric properties. This results in edge-preserved image smoothing, which is the purpose of the bilateral filter. It should be noted, however, that the foregoing equations are mathematical representations for filtering of infinite-precision images. As infinite-precision is a concept more pertinent to the analog domain and does not exist in the digital domain, each double integral in equations 1-6 must be understood as the summation of a finite set of calculations, i.e. one calculation per Neighbor and one set of calculations per Center Pixel, or put another way, one set of calculations per Filter iteration. A formal expression is given below:
\begin{equation}
\label{eq:depen}
filtered\: pixel\: x=\frac{\substack{N\\
\sum\\
i=1
}
f(\xi_{i})\cdot c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}{\substack{N\\
\sum\\
i=1
}
c(\xi_{i},x)\cdot s(f(\xi_{i}),f(x))}
\end{equation}

In addition to restating the bilateral filter's mathematical model in terms of a finite summation, Equation~\ref{eq:depen} reveals the data dependency that exists in the bilateral filter. Specifically, Equation~\ref{eq:depen} shows that before the filtered pixel can be calculated by dividing the numerator by the denominator, both the summation in the numerator and denominator must first be calculated. Consequently, the bilateral filter has two data dependencies. As explained later in this paper, the pair-symmetric algorithm significantly increases the memory consumption in order to manage data dependencies but pays off by cutting, in half, the total number of executions.

Another observation taken into account is that the values of Equation~\ref{spatial}, for the Neighborhood, can be computed in advance as the values depend on a simple Gaussian probability distribution function. However, the values of Equation~\ref{photometric} and consequently of Equation~\ref{combined} cannot be known a priori as they depend on the similarity of signals between the current Center Pixel and the Neighborhood. Moreover, as the size of the Window increases, which only happens by choice of the user as the size is user defined, the number of computations per pixel grows at a rate of $O(FilterWindowWidth^{2})$ where FilterWindowWidth is the diameter of the Window. By reason of the foregoing, the bulk of the time spent in processing an image using the bilateral filter is in the arithmetic-intensive computations, because the number of writes to the destination image is constant as the rate of growth of Center-Neighbor computations with respect to the Window size increase.

\subsection{A Naive Approach to bilateral filtering kernel}

\begin{figure}
\centering\includegraphics[scale=0.35]{images/the1}
\includegraphics[clip,scale=0.35]{images/the202}\includegraphics[scale=0.35]{images/the2}\includegraphics[scale=0.35]{images/the3}
\caption{Example of a naive bilateral filter iterations}
\end{figure}

The naive stencil algorithm is the standard model that is predominantly used in bilateral filtering kernel. Previous implementations\cite{Paris2009,Pham2005} of bilateral filtering kernel improves the hardware dependent aspects and hence expend resources on achieving performance gains that can not match the effort taken. Here, we address a root problem of the standard naive algorithm by proposing a new algorithm, and then develop hardware dependent improvements for further speedup. Algorithm~\ref{alg:Example-Algorithm-float} is the pseudo-code for the standard naive bilateral filtering kernel.

\begin{algorithm}[!h]
\caption{Naive bilateral Filter Algorithm}
\label{alg:Example-Algorithm-float}
\begin{algorithmic}
\FORALL{element  $s_{i,j}$ in source image s}
\STATE{$X_{1} \leftarrow 0$}
\STATE{$X_{2} \leftarrow 0$}
\FOR{$k=-filter\_hw$ \TO $filter\_hw$}
\FOR {$l  = -filter\_hw$ \TO $filter\_hw$}
\STATE{$temp \leftarrow e^{C_1(k^2 + l^2) + C_2(s[i,j] - s[i,j+l])^{2}}$}
\STATE{$X_1 \leftarrow X_1 + temp \cdot s[i+k,j+l]$}
\STATE{$X_2 \leftarrow X_2 + temp$}
\ENDFOR
\ENDFOR
\STATE{output[$i,j] \leftarrow X_{1} / X_{2}$}
\ENDFOR
\end{algorithmic}
\end{algorithm}

\subsection{Pair-symmetric bilateral filtering kernel}
The pair-symmetric algorithm, shown in Algorithm~\ref{alg:Example-Algorithm-float-1}, reduces the amount of Center-Neighbor computations by half. Thus, it can theoretically cut the processing time in half at the cost of increased memory requirement, which increases with the amount of dependencies. Its implementation is non-trivial due to potential race-conditions.

Our pair-symmetric bilateral filtering algorithm is premised on the fact that each pixel assumes the role of both a Center Pixel and a Neighboring Pixel, and that the codes are symmetric with respect to Center-Neighbor paired computations. For example, in the bilateral filter, $c(\xi,x)$ returns the same value as $c(x,\xi)$ and $s(f(\xi),f(x))$ returns the same value as $s(f(x),f(\xi))$. It then follows that the naive bilateral filtering algorithm dictated in Algorithm~\ref{alg:Example-Algorithm-float} above, although simple, performs twice as many calculations than necessary.

The Pair-symmetric algorithm eliminates this redundancy that is inherent in the naive bilateral filtering algorithm. Instead of reading from all Neighbors, our pair-symmetric algorithm reads from only half of the Neighbors. The result of each Center-Neighbor paired computation is then added to the $X_n$ register, which eventually obtains the result of data dependency $n$. This is trivial as the naive bilateral filtering algorithm already does this. What our pair-symmetric algorithm does unique is that the result of each Center-Neighbor paired computation is also added to and stored in the Neighbor's position in a temporary array ("Map'') that is used for storing the results of a data dependency. So, there are $n$ Maps for $n$ number of dependencies.

After the data dependency conditions are met, the function that returns the output data from the dependencies iterates over the Maps. As applied to the bilateral filter, the data dependency lies in calculating each pixel's numerator and denominator that is used to derive the pixel's output result. After the numerators and denominators are derived, a function iterates and divides the corresponding elements of the two arrays to return the output.

@@

\section{CUDA-based Implementation Details}
\label{sec:GPUoptimizations} 
As bilateral filtering algorithm belongs to the class of embarrassingly parallel algorithms, CUDA is well suited for its implementation. However,
in order to effectively implement such an algorithm in CUDA, one must understand how CUDA operates, which is significantly different from any CPU.

@@

In CUDA, threads are organized into groups called warps, and warps are launched in groups called thread-blocks, in which each thread-block can share among its warps an allocated portion of shared-memory. Each warp executes one instruction at a time for its own group of threads. Because all threads in a warp share the same program counter and are therefore driven by the same sequential instruction stream, all threads in a warp are always in sync with each other without requiring explicit synchronization calls. Therefore, synchronization between threads in a warp is implicit.

To take full advantage of CUDA's implicit intra-warp synchronization, divergent branching between threads in the same warp must be avoided. This is so because all the threads in the same warp share only one program counter. A divergent branch would cause some or most of the threads to be idle as the program counter would run through the branch and drive only those threads that met the branch condition to work. An example of a divergent branch can be a simple if-statement such as: $if~(threadIdx.x < 5)~\{...\}~else~\{...\}$.

Divergent branching commonly occurs in memory access patterns in which global memory operations are not coalesced and shared memory operations are not free of memory bank conflicts. For example, if eight threads were to read from eight non-contiguous segments of global memory---commonly seen in strided access patterns like array {[}threadIdx.x {*} 2{]}---divergent branching would occur because a warp of threads can only read from one segment of memory per instructional transaction, and that segment's word-length is hardware defined. In other words, a memory transaction would be serialized if all the threads in the warp to which the transaction belonged were not aligned consecutively or did not fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For the Bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*width + j] representation of a 2D image, for if threads in a warp were to read from or write to array elements (representing pixels) in a vertical order, there would be a separate transaction for each thread in the warp.

In other words, a memory transaction will be serialized unless all the threads in the warp to which the transaction belongs are aligned consecutively and fit within a segment of memory. Therefore, in order for all the threads of a warp to read/write from global memory in a single transaction, the threads must correspond read/write operations to aligned contiguous addresses within a segment of memory. For bilateral filtering algorithm, warps should correspond to horizontal rows of pixels for an [i*jMax + j] representation of a 2D image, for if threads in a warp were to operate on pixels in a vertical order, there would be a separate transaction for each thread in the warp.

While intra-warp synchronization is guaranteed, the rate at which warps stream through their sequence of instructions is unpredictable. As a consequence, work must be arranged such that warps can work as independent from each other as possible, in order to minimize explicit synchronization calls. This pooling of work is achieved by threads reading from addresses incrementally along the horizontal axis, and only making an explicit synchronization call per vertical shift to the next row once the work requiring the warp's access to the current row of input data has completed. This design works because horizontal shifts are implicitly synchronized as there is only one warp per row, but a vertical shift without an explicit synchronization call will result in a warp trespassing into the territory of another warp, thereby causing race conditions as threads from two different warps will consequently be prone to write to the same memory location.
